Modal Logics in the Coq Proof Assistant

### 2014-12-09 Papers/InProgress/ModalLogicCoq/ModalCoq.tex
Write section 3 (tactics for modalities)

@@

\begin{calculus}
{propositional rules}
{fig:PropositionalRules}

@@

\s\s
\infer[\wedge_I]{A \wedge B}{A & B}
\s\s
\infer[\wedge_{E_1}]{A}{A \wedge B}
\s\s
\infer[\wedge_{E_2}]{B}{A \wedge B}

% \vspace{2em}

% \s\s
% \infer[\vee_E]{C}{A \vee B & \infer*{C}{\infer{A}{}} & \infer*{C}{\infer{B}{}}}
% \s\s
% \infer[\vee_{I_1}]{A \vee B}{A}
% \s\s
% \infer[\vee_{I_2}]{A \vee B}{B}

@@

\begin{calculus}
{double negation elimination}
{fig:Classical}
\infer[\neg\neg_E]{A}{ \neg\neg A }
\end{calculus}

\begin{calculus}
{quantifier rules}
{fig:QuantifierRules}

\s
\infer[\all_I]{\all x_{\tau}. A[x]}{ A[\alpha] }
\s
\infer[\all_E]{A[t]}{ \all x_{\tau}. A[x] }
\s\s
\infer[\ex_I]{\ex x_{\tau}. A[x]}{ A[t] }
\s
\infer[\ex_E]{A[\beta]}{ \ex x_{\tau}. A[x] }

\begin{center}
\textbf{eigen-variable conditions:} \\
if $\rho$ is a $\all_I$ inference eliminating a variable $\alpha$, then any occurrence of $\alpha$ in the proof should be an ancestor of the occurrence of $\alpha$ eliminated by $\rho$; \\
if $\rho$ is a $\ex_E$ inference introducing a variable $\beta$, then any occurrence of $\beta$ in the proof should be a descendant of the occurrence of $\beta$ introduced by $\rho$.
\end{center}

\end{calculus}

@@

\begin{center}
\textbf{eigen-box condition:}\\ 
$\omega$ must be a fresh name for a box. \\
\vspace{0.5em}
\textbf{boxed assumption condition:} \\
assumptions should be discharged within the box where they are made.
\end{center}

\end{calculus}

\noindent
To maximally preserve user intuition in interactive modal logic theorem
proving, the embedding via the possible world semantics should be as
transparent as possible to the user. Fortunately, basic Coq tactics
such as \texttt{intro}, \texttt{apply} and \texttt{split} automatically
unfold the modal definitions. They can thus be used with modal
connectives and quantifiers just as they are used with the usual
connectives and quantifiers.  For the modal operators, however, the
following simple tactics have been implemented, in order to allow the
user to work with the concepts of \emph{necessity} and
\emph{possibility} without having to unfold the definitions of modal
operators and think in terms of possible worlds.

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Ltac box_i := \llet w := fresh "w" \iin \llet R := fresh "R" 
               \iin (intro w at top; intro R at top).

\Ltac box_elim H w1 H1 := \match type of H \with 
      ((box ?p) ?w) => cut (p w1); 
                       [intros H1 | (apply (H w1); try assumption)] \eend.

\Ltac box_e H H1:= \match goal \with | [ |- (_ ?w) ] => box_elim H w H1 \eend.

\Ltac dia_e H := \llet w := fresh "w" \iin \llet R := fresh "R" \iin 
                (destruct H \as [w [R H]]; move w at top; move R at top).

\Ltac dia_i w := (exists w; split; [assumption | idtac]).
\end{Verbatim}

\noindent
Validity of a modal proposition is validity of its grounding on any world. A non-intrusive notation and a tactic that grounds the proposition to an arbitrary world are provided.

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Definition V (p: o) := \fforall w, p w.
\Notation "[ p ]" := (V p).
\Ltac mv := \match goal \with [|- (V _)] => intro \eend.
\end{Verbatim}

>>>>

\noindent
A modal proposition is valid iff it holds in every possible world. 
This notion of modal validity is encoded by the following defined predicate:

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Definition V (p: o) := \fforall w, p w.
\end{Verbatim}

\noindent 
To prove a modal proposition \texttt{p} (of type \texttt{o})
within \Coq, the proposition \texttt{(V p)} (of type \texttt{Prop})
should be proved instead. To increase the transparency of the
embedding to the user, the following notation is provided, allowing
\texttt{[ p ]} to be written instead of \texttt{(V p)}.

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Notation "[ p ]" := (V p).
\end{Verbatim}

\section{Tactics for Modalities}

Interactive theorem proving in \Coq is usually done with tactics,
imperative commands that reduce the theorem to be proven (i.e. the
goal) to simpler subgoals, in a bottom-up manner. The simplest tactics
can be regarded as rules of a natural deduction calculus\footnote{The
underlying proof system of \Coq (the Calculus of Inductive
Constructions \cite{Paulin}) is actually more sophisticated and
minimalistic than the calculus shown in Figure
\ref{fig:NaturalDeduction}. But the calculus shown here suffices for
the purposes of this paper.} (e.g. as those shown in Figure
\ref{fig:NaturalDeduction}).  For example: the \texttt{intro} tactic
can be used to apply the introduction rules for implication and for
the universal quantifier; the \texttt{apply} tactic corresponds to the
elimination rules for implication and for the universal quantifier;
\texttt{split} performs conjunction introduction; \texttt{exists} can
be used for existential quantifier introduction and \texttt{destruct}
for its elimination.

@@

\begin{calculus}
{An arbitrary natural deduction calculus}
{fig:NaturalDeduction}
\begin{small}

@@

\s
\infer[\neg\neg_E]{A}{ \neg\neg A }
\s\s
\infer[\wedge_I]{A \wedge B}{A & B}
\s\s
\infer[\wedge_{E_1}]{A}{A \wedge B}
\s\s
\infer[\wedge_{E_2}]{B}{A \wedge B}

\s\s
\infer[\vee_E]{C}{A \vee B & \infer*{C}{\infer{A}{}} & \infer*{C}{\infer{B}{}}}
\s\s
\infer[\vee_{I_1}]{A \vee B}{A}
\s\s
\infer[\vee_{I_2}]{A \vee B}{B}

%\begin{scriptsize}

\s\s\s
\infer[\all_I]{\all x_{\tau}. A[x]}{ A[\alpha] } 
\s\s
\infer[\all_E]{A[t]}{ \all x_{\tau}. A[x] }

\s\s\s
\infer[\ex_I]{\ex x_{\tau}. A[x]}{ A[t] }
\s\s
\infer[\ex_E]{C}{ \ex x_{\tau}. A[x] & \infer*{C}{A[\alpha]} }

%\end{scriptsize}

\begin{center}
$\alpha$ must respect the usual \emph{eigen-variable conditions}.
\end{center}

\end{small}

@@

To maximally preserve user intuition in interactive modal logic
theorem proving, the embedding via the possible world semantics should
be as transparent as possible to the user. Fortunately, the basic Coq
tactics described above automatically unfold the shallowest modal
definition in the goal.  Therefore, they can be used with modal
connectives and quantifiers just as they are used with the usual
connectives and quantifiers. The situation for the new modal
operators, on the other hand, is not as simple, unfortunately.

Since the modal operators are, in our embedding, essentially just
abbreviations for quantifiers guarded by reachability conditions, the
typical tactics for quantifiers can be used, in principle.  However,
this exposes the user to the technicalities of the embedding,
requiring him to deal with possible worlds and their reachability
explicitly. In order to obtain transparency also for the modal
operators, we have implemented a few specialized tactics using \Coq's
Ltac language.

When applied to a goal of the form \texttt{((box p) w0)}, the tactic
\texttt{box\_i} will introduce a fresh new world \texttt{w} and then
introduce the assumption that \texttt{w} is reachable from
\texttt{w0}. The new goal will be \texttt{(p w)}.

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Ltac box_i := \llet w := fresh "w" \iin \llet R := fresh "R" 
               \iin (intro w at top; intro R at top).
\end{Verbatim}

\noindent 
If the hypothesis \texttt{H} is of the form \texttt{((box p)
w0)} and the goal is of the form \texttt{(q w)}, the tactic
\texttt{box\_e H H1} creates a new hypothesis \texttt{H1: (p w)}. The
tactic \texttt{box\_elim H w1 H1} is an auxiliary tactic for
\texttt{box\_e}. It creates a new hypothesis \texttt{H1: (p w1)}, for
any given world \texttt{w1}, not necessarily the goal's world
\texttt{w}. It is also responsible for automatically trying (by
\texttt{assumption}) to solve the reachability guard conditions,
releasing the user from this burden.

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Ltac box_elim H w1 H1 := \match type of H \with 
      ((box ?p) ?w) => cut (p w1); 
                       [intros H1 | (apply (H w1); try assumption)] \eend.
\end{Verbatim}

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Ltac box_e H H1:= \match goal \with | [ |- (_ ?w) ] => box_elim H w H1 \eend.
\end{Verbatim}

\noindent 
If the hypothesis \texttt{H} is of the form \texttt{((dia p)
w0)}, the tactic \texttt{dia\_e H} generates a new hypothesis
\texttt{H: (p w)} for a fresh new world \texttt{w} reachable from
\texttt{w0}.

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Ltac dia_e H := \llet w := fresh "w" \iin \llet R := fresh "R" \iin 
                (destruct H \as [w [R H]]; move w at top; move R at top).
\end{Verbatim}

\noindent
The tactic \texttt{dia\_i w} transforms a goal of the form 
\texttt{((dia p) w0)} into the simpler goal \texttt{(p w)} 
and automatically tries to solve the guard condition that 
\texttt{w} must be reachable from \texttt{w0}.

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Ltac dia_i w := (exists w; split; [assumption | idtac]).
\end{Verbatim}

\noindent If the new modal tactics above are regarded from a natural
deduction point of view, they give rise to the inference
rules\footnote{The natural deduction calculus with the rules from
Figures \ref{fig:NaturalDeduction} and \ref{fig:NDK} is sound and
complete relatively to the calculus of Figure
\ref{fig:NaturalDeduction} extended with a necessitation rule and the
modal axiom K \cite{ToDo:UnpublishedPaperInGithub}. Therefore,
assuming that the calculus from \ref{fig:NaturalDeduction} is sound
and Henkin-complete for classical higher-order logic, the additional
modal rules in Figure \ref{fig:NDK} make it sound and Henkin-complete
for modal higher-order logic \textbf{K}.} shown in Figure
\ref{fig:NDK}. The labels that name boxes in those inference rules are
precisely the worlds that annotate goals and hypotheses in \Coq with
the modal embedding. A hypothesis of the form \texttt{(p w)}, where
\texttt{p} is a modal proposition of type \texttt{o} and \texttt{w} is
a world of type \texttt{i} indicates that \texttt{p} is an assumption
created inside a box with name \texttt{w}.

@@

\begin{center}
\textbf{eigen-box condition:}\\ 
$\omega$ must be a fresh name for a box. \\
\vspace{0.5em}
\textbf{boxed assumption condition:} \\
assumptions should be discharged within the box where they are created.
\end{center}

Finally, our implementation also provides the tactic \texttt{mv},
standing for \emph{modal validity}, which replaces a goal of the form
\texttt{[ p ]} (or equivalently \texttt{(V p)}) by a goal of the form
\texttt{(p w)} for a fresh arbitrary world \texttt{w}.

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Ltac mv := \match goal \with [|- (V _)] => intro \eend.
\end{Verbatim}

\section{Convenient Modal Lemmas}
